---
title: "DATA607 Final Project"
author: "IvanTikhonov"
date: "2022-11-29"
output: html_document
---


Introduction

Goal
I seek to understand the rates of emergency response incidents and then compare this 
with macro factors such as potentially preventable emergency visits by patient county and financial districts health,
as well as Covid-19 data to determine if there is a correlation.




Motivation for performing the analysis
My wife works as a nurse in the Emergency Room at one of the NY hospitals. 
I want to conduct this analysis to help her determine which county has the most favorable conditions for potential employment.


Data Details

 Data for this analysis is made up of four components.

 The first dataset represents All Payer Potentially Preventable Emergency Visit Rates by Patient County from NY Health Data 
 https://health.data.ny.gov/Health/All-Payer-Potentially-Preventable-Emergency-Visit-/f8ue-xzy3


 The second dataset represents Local Area Unemployment Statistics
 https://data.ny.gov/Economic-Development/Local-Area-Unemployment-Statistics-Beginning-1976/5hyu-bdh8/data

 The third dataset represents Neighborhood Financial Health Digital Mapping and Data Tool from NYC Open Data.
 https://data.cityofnewyork.us/Business/Neighborhood-Financial-Health-Digital-Mapping-and-/r3dx-pew9
 
 The fourth dataset represents Covid-19 rates across boroughs. 
 https://github.com/nychealth/coronavirus-data
 

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

Load library
```{r}
library(tidyverse)
library(rlang)
library(purrr)
library(magrittr)
library(ggplot2)
library(corrplot)
library(xfun)
library(kableExtra)
library(sda)
library(entropy)
library(corpcor)
library(fdrtool)
library(tidyr)
library(dplyr)
library(zip)
library(st)
library(rlang)
library(set)
library(wk)
library(s2)
library(crs)
```


Load Data
```{r}
# Raw csv Url for All Payer Potentially Preventable Emergency Visit Rates by Patient County from NY Health Data.
EDVisit <- read.csv( "https://raw.githubusercontent.com/IvanGrozny88/DATA607-Final-Project/main/All_Payer_Potentially_Preventable_Emergency_Visit__PPV__Rates_by_Patient_County__SPARCS____Beginning_2011_Chart.csv", sep=",",header = TRUE)
unemployment<-read.csv("https://raw.githubusercontent.com/IvanGrozny88/DATA607-Final-Project/main/Local_Area_Unemployment_Statistics__Beginning_1976.csv", sep=",",header = TRUE)
NFHDM<-read.csv("https://raw.githubusercontent.com/IvanGrozny88/DATA607-Final-Project/main/Neighborhood_Financial_Health_Digital_Mapping_and_Data_Tool.csv", sep=",",header = TRUE)
covid<-read.csv("https://raw.githubusercontent.com/nychealth/coronavirus-data/master/latest/pp-by-modzcta.csv", sep=",",header = TRUE)
```


In this section, I have listed our given data into yearly results, with hopes of finding some patterns; two of these patterns could be related to the borough and yearly emergency visits, for example.


```{r}
EDVisit %>%
  group_by(Discharge.Year) %>%
  top_n(15) %>%
  ungroup() %>%
  mutate(Patient.County.Name = reorder(Patient.County.Name, Difference.in.Rates)) %>%
  ggplot(aes(Patient.County.Name, Difference.in.Rates, fill = Discharge.Year)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Discharge.Year, scales = "free_y") +
  labs(y = "Emergency Visit",
       x = NULL) +
  coord_flip()
```








we can quickly observe a few interesting patterns of observed emergency discharges summary yearly Values:
```{r}
# we are generatin colum for each yearly Visit
EDVisit$num_Expected.Rate.Per.100.People <- 1
TotalRate.Per.100.People <- EDVisit$num_Expected.Rate.Per.100.People %>% sum()
Discharge.Year.data <- EDVisit %>%
                    group_by(Discharge.Year) %>%
                    summarise(`Observed.Rate.Per.100.People` = sum(num_Expected.Rate.Per.100.People),
                               Percentage = paste(round((`Observed.Rate.Per.100.People`/TotalRate.Per.100.People)*100,2),"%"))  %>%
                    arrange(desc(`Discharge.Year`))
kable(Discharge.Year.data)
```
Below we discover interesting patterns of observed yearly emergency visits Values by the borough:

```{r}
boroughPatient.County.Name.data <- EDVisit %>%
                    group_by(Patient.County.Name) %>%
                    summarise(`Observed.Rate.Per.100.People` = sum(num_Expected.Rate.Per.100.People),
                               Percentage = paste(round((`Observed.Rate.Per.100.People`/TotalRate.Per.100.People)*100,2),"%") )  %>%
                    arrange(desc(`Patient.County.Name`))
kable(boroughPatient.County.Name.data)
```
Procedure to Find summary daily Values by the borough


```{r}
boroughPatient.County.Name.data <- EDVisit %>%
                    group_by(Discharge.Year, Risk.Adjusted.Rate.Per.100.People, Expected.Rate.Per.100.People, Patient.County.Name) %>%
                    summarise(`Observed.Rate.Per.100.People` = sum(num_Expected.Rate.Per.100.People))  %>%
                    arrange(desc(`Risk.Adjusted.Rate.Per.100.People`))
                    
boroughPatient.County.Name.data <- spread(data = boroughPatient.County.Name.data,
                             key = Patient.County.Name,
                             value = `Observed.Rate.Per.100.People`
                             )

colnames(boroughPatient.County.Name.data)[9] <- "Not Available"
kable(head(boroughPatient.County.Name.data,10))
```


To make the above datasets user-friendly, I transformed the columns into rows:



```{r}
# transform the columns in to rows data
boroughPatient <- gather(boroughPatient.County.Name.data, "borough", "Patient", 4:65)
boroughPatient
```









The next step was to compare expected rates with risk adjusted rates to better analyze the data:



```{r}
bind_rows(boroughPatient, 
          boroughPatient.County.Name.data) %>%
  ggplot(aes(Risk.Adjusted.Rate.Per.100.People, Expected.Rate.Per.100.People, fill = Discharge.Year)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Discharge.Year, ncol = 1, scales = "free_y")
```

Furthermore, I wanted to find out how Covid-19 affected various zip codes and the rates of the disease. In order to analyze that, I cleaned up the zip code variable by removing extra characters.

```{r}
# filtered days, made the dataset longer by putting zip codes in a column
covid <- covid %>%
  filter(End.date == "11/27/2022") %>%
  pivot_longer(2:184, names_to = "zipcode", values_to = "rate") %>%
  set_colnames(c("date", "zipcode", "rate"))

# cleaned up the zip code variable by removing extra characters
covid$zipcode <- covid$zipcode %>%
  str_remove("X") %>%
  str_replace("\\.", " ")

zipcodes <- covid[(7:183), ] %>%
  dplyr::select(zipcode) %>%
  as.vector()
covid 
```

Below is the graphical representation of these databases:





```{r}
bind_rows(covid) %>%
  ggplot(aes(zipcode, rate, fill = date)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~date, ncol = 1, scales = "free_y")
```




Another important step was to compare expected rates with observed rates to better analyze the data:


```{r}
ggplot(EDVisit, aes(Observed.Rate.Per.100.People, Expected.Rate.Per.100.People, fill = Discharge.Year)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Discharge.Year, ncol = 2, scales = "free_x")

```


To better understand the importance of topic, I included year unemployed percentage and financial districts health. 



```{r}
unemployment$num_Labor.Force <- 1
Total <- unemployment$num_Labor.Force %>% sum()
Year.data <- unemployment %>%
                    group_by(Year) %>%
                    summarise(`Unemployed` = sum(num_Labor.Force),
                               Percentage = paste(round((`Unemployed`/Total)*100,2),"%"))  %>%
                    arrange(desc(`Year`))
kable(Year.data)
```



Another valuable task was to include medial income by the borough:




```{r}
NFHDM %>%
  group_by(Year.Published) %>%
  top_n(5) %>%
  ungroup() %>%
  mutate(Borough = reorder(Borough, Median_Income)) %>%
  ggplot(aes(Borough, Median_Income, fill = Year.Published)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Year.Published, scales = "free_y") +
  labs(y = "Median_Income by Borough ",
       x = NULL) +
  coord_flip()
```


Conclusion and Recommendation:


After modeling analysis, there is a statistically significant linear relationship between the potentially preventable
emergency visits by patient county and financial districts health, including unemployment statistics. 
 A high-level median income within all of NYC (inclusive of all five boroughs- Queens, Manhattan, Bronx, Staten Island, Brooklyn) displayed that Brooklyn and Manhattan are the two dominating boroughs. Unfortunately, due to the high data variability, I could not relate the patterns to the borough and yearly emergency visits.
The unemployed percentage is broken out by year the highest number of cases:
2022: decrease in cases
2010- 2021: steady number of cases with a slight increase
2000-2009: steady number of cases
 Overall, Manhattan has the lowest Covid-19 rates among the five boroughs. As a result,
 I recommend my wife the following borough as-top preferred workplace due
 to the higher number of employed individuals and better financial district health.











